Visualisations for RSE Curious event

### 2023-07-13 visualisations_for_rse_curious_event.qmd
made new script for the RSE curious event, to maek a simple visualisation

---
title: "Visualisations for short paper, Feb 2023"
author: "Jan savinc"
format: html
editor_options: 
  chunk_output_type: console
---

@@

```{r}
dir_paper <- file.path(dir_outputs, "short_paper_feb2023")
if (!dir.exists(dir_paper)) dir.create(dir_paper)
```

@@

# Place of death image

## Wrangle data

```{r}
merged_deaths_overall_2015_2019_historic_period <-
  bind_rows(
    ## pre-2022 data
    merged_deaths_overall %>% filter(year < 2022),
    ## 2022 data
    merged_deaths_overall %>%
      filter(year >= 2022) %>%  # 2022 or later!?
      select(-matches("deaths\\_mean|deaths\\_min|deaths\\_max")) %>%  # remove data using NRS historic period
      mutate(historic_period = "2015-2019") %>%  # join the previous hsitoric period data instead
      left_join(historical_weekly_average_overall, by = c("historic_period","week_number","place_of_death")) %>%
      compute_moving_average_for_current_data(  # not strictly required here but will be easier to apply in other places if included
        current_data = ., 
        historical_data = 
          weekly_deaths_2015_2019_overall %>% mutate(year=as.integer(year)) %>% drop_na(year),  # remove the entries with 'average' for year
        num_weeks_to_average = 4)
  )
```

## With 2015-2019 historic range only

Note that the footnote to the image should be separate to the file so I'll remove the title and captions and export them as a text file instead.

```{r}
fig_paper_pod_2015_2019_range <-
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(!place_of_death %in% c("All","Other")) %>%
  mutate(place_of_death = factor(place_of_death, levels = order_of_place_of_death_levels[c(1,3,2)])) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non-Covid deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)"
       ) +
  ## legend to include historic range & mean
  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
  NULL

fig_paper_pod_2015_2019_range %>%
save_output_file(filename = file.path(dir_paper, paste0(c("fig_paper_pod_2015_2019_range", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")

fig_paper_pod_2015_2019_range_footnote <- c(
  date_range_text,
  paste0(c(death_registrations_text,historic_period_2015_2019_text,source_nrs_text), collapse="\n")
)

fig_paper_pod_2015_2019_range_footnote %>%
  write_lines(file = file.path(dir_paper, paste0("fig_paper_pod_2015_2019_range_", latest_data_modified_date, ".txt")))
```

### As above, but split into 1a,b,c

```{r}
p1 <- 
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non-Covid deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)"
       ) +
  ## legend to include historic range & mean
  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
  NULL

p2 <-
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Hospital")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non-Covid deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = NULL
       ) +
  ## legend to include historic range & mean
  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
  NULL

p3 <-
  merged_deaths_overall_2015_2019_historic_period %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Care home")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = date_w_c)) +
  add_vertical_lines_and_annotations_for_years(years = 2021:current_year, y_annotation = 30) +
  scale_x_date(date_labels = "%d-%m-%y", date_breaks = "12 weeks", limits = c(earliest_date, most_recent_date_available_2023)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max, fill = "grey70"), alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean, linetype = "dashed"), colour="grey20") +
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non-Covid deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(5, 15, 5, 5),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = NULL
       ) +
  ## legend to include historic range & mean
  scale_linetype_manual(values = c("dashed"), labels = "Historic weekly mean") +
  scale_fill_manual(values = c("grey50"), labels = "Historic weekly range") +
  NULL

fig_paper_pod_2015_2019_range_1abc <-
  p1 + (p2 / p3) +
    plot_layout( 
      guides = "collect",
      widths = c(2,1)
    ) +
    plot_annotation(tag_levels = "a", tag_suffix = ")") &
    # if (interactive()) theme_minimal() else NULL &  # for testing
    theme(
      legend.position = "bottom", 
      plot.tag = element_text(size = rel(1.1))
      )

fig_paper_pod_2015_2019_range_1abc %>%
save_output_file(filename = file.path(dir_paper, paste0(c("fig_paper_pod_2015_2019_range_abc_layout", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png",".tiff"), plot = ., width = 28, height = 14, dpi = 300, units = "cm")

fig_paper_pod_2015_2019_range_1abc_footnote <- c(
  "a) Home & other non-insitution, b) Hospital, c) Care home; 'Other' place not shown due to small numbers.",
  fig_paper_pod_2015_2019_range_footnote
)

fig_paper_pod_2015_2019_range_1abc_footnote %>%
  write_lines(file = file.path(dir_paper, paste0("fig_paper_pod_2015_2019_range_1abc_footnote_", latest_data_modified_date, ".txt")))
```

# Place of death by LA

```{r}
fig_proportion_of_deaths_at_home_by_la <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "la") %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
  pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
  ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
  geom_point() +
  geom_line() +
  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
  scale_linewidth_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  scale_x_continuous(breaks = sort(unique(proportions_of_deaths_by_place_by_geography$year))) +
  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12),
    plot.caption = element_text(size = 10, colour = "gray60"),
    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
    strip.background = element_rect(fill = NA, colour = "gray80"),
    axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
  labs(
    linetype = NULL,
    x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
    subtitle = paste0("* ",only_2023_disaggregated_text),
    # caption = paste0(c("Local Authorities ordered by 2020 population estimate.",death_registrations_text,source_nrs_text),collapse="\n")
    ) +
  NULL

fig_proportion_of_deaths_at_home_by_la %>%
save_output_file(filename = file.path(dir_paper, paste0(c("fig_proportion_of_deaths_at_home_by_la", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png", ".tiff"), plot = ., width = 30, height = 25, dpi = 300, units = "cm")

fig_proportion_of_deaths_at_home_by_la_footnote <- c(
  "Local Authorities are shown in order of 2020 popoulation estimate.",
  date_range_text_disaggregated,
  paste0(c(death_registrations_text,source_nrs_text), collapse="\n")
)
fig_proportion_of_deaths_at_home_by_la_footnote %>%
  write_lines(file = file.path(dir_paper, paste0("fig_proportion_of_deaths_at_home_by_la_footnote_", latest_data_modified_date, ".txt")))
```

>>>>

---
title: "Visualisations for RSE Curious event"
author: "Jan Savinc"
format: html
editor_options: 
  chunk_output_type: console
---

@@

```{r}
dir_event <- file.path(dir_outputs, "rse_curious_2023")
if (!dir.exists(dir_event)) dir.create(dir_event)
```

# Download & process monthly data

This section maybe belongs in the earlier pipeline?

NRS publish monthly mortality reports broken down by various variables, including Place of death over years, available since 2005.

These are updated monthly, and reported for the previous month, so as of 12 July 2023, the latest report was published in June, and contains data for up to May 2023.]

Note also that these exhibit small differences from the weekly data - the weekly data are reported to the ISO 8061 standard, meaning that week 1 of 2020 starts in 2019, for example.

```{r}
monthly_report_url <- "https://www.nrscotland.gov.uk/files//statistics/vital-events/monthly-mortality/monthly-mortality-may-23-tab-5.xlsx"

temporary_report_path <- tempfile()
download.file(url = monthly_report_url, destfile = temporary_report_path, mode = "wb")

# excel_sheets(temporary_report_path)
# deaths_monthly <- read_excel(temporary_report_path, sheet = 5)  # more reliable to use name, though that is liable to change
deaths_monthly <- 
  read_excel(temporary_report_path, sheet = "Table M5", skip = 4) %>%
  clean_names() %>%
  rename(
    year = year_of_occurrence,
    month = month_of_occurrence,
    place_of_death = location
  ) %>%
  select(-provisional_data_marker) %>%
  mutate(
    place_of_death = order_of_place_of_death_levels[match(x = str_sub(place_of_death,1,4), table = str_sub(order_of_place_of_death_levels, 1, 4))] %>%
      factor(., levels = order_of_place_of_death_levels)
  )
```

@@

# Monthly place of death image

Monthly deaths might be too detailed...

```{r}
deaths_monthly %>%
  mutate(
    date = ymd(glue::glue("{year}-{month}-01"))
  ) %>%
  filter(place_of_death == "Home & other non-institution") %>%
  filter(year >= 2015) %>%
  # ggplot(aes(x = paste(year,month,sep="-"), number_of_deaths)) +
  ggplot(aes(x = date, y = number_of_deaths)) +
  add_vertical_lines_and_annotations_for_years(years = 2015:current_year, y_annotation = 30) +
  geom_line() +
  # facet_grid(place_of_death~.) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1)
  ) +
  labs(x = NULL,
       y = "Deaths (N)"
       ) +
  NULL
```

# Annual home deaths

```{r}
deaths_annual <-
  deaths_monthly %>%
  group_by(year, place_of_death) %>%
  summarise(
    number_of_deaths = sum(number_of_deaths),
    .groups = "drop"
  )

annual_deaths_by_place_incl_covid <-
  bind_rows(
    ## 2015-2019
    merged_proportions_of_deaths_by_place %>%
    select(year, place_of_death, number_of_deaths) %>%
    filter(year %in% 2015:2019) %>%
    mutate(deaths_non_covid = number_of_deaths, deaths_covid_related = 0)
    ,
    ## 2020 onwards
    total_deaths_overall %>%
      select(year, place_of_death, number_of_deaths = deaths_all_causes, deaths_covid_related) %>%
      distinct %>%
      mutate(deaths_non_covid = number_of_deaths - deaths_covid_related)
  )

# merged_proportions_of_deaths_by_place %>%
#   select(year, place_of_death, number_of_deaths_from_weekly = number_of_deaths) %>%
#   left_join(deaths_annual, by = c("year","place_of_death")) %>%
#   filter(place_of_death!="All") %>%
#   ggplot(aes(x=year, colour = place_of_death)) +
#   geom_line(aes(y = number_of_deaths), linetype = "solid") +
#   geom_point(aes(y = number_of_deaths), shape = 10) +
#   geom_line(aes(y = number_of_deaths_from_weekly), linetype = "dashed") +
#   geom_point(aes(y = number_of_deaths_from_weekly), shape = 15) +
#   NULL

(fig_annual_home_deaths <- 
  annual_deaths_by_place_incl_covid %>%
  filter(place_of_death == "Home & other non-institution") %>%
  filter(year >= 2015 & year != 2023) %>%
  pivot_longer(cols = c(deaths_covid_related, deaths_non_covid), names_to = "death_type", values_to = "deaths") %>%
  mutate(bin_pandemic = year >= 2020) %>%
  ggplot(aes(x = year, y = deaths, fill = death_type)) +
  geom_col() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 60, colour = "gray40", vjust=1, hjust=1), 
    aspect.ratio = 0.5
  ) +
  scale_x_continuous(breaks = 2015:2022) +
  scale_y_continuous(labels = scales::comma) +
  # scale_fill_manual(values = scales::viridis_pal(option = "D")(3)[c(3,1)],labels = c("Covid-related", "Non-covid")) +
  scale_fill_manual(values = c("red","grey50"),labels = c("Covid-related", "Non-covid")) +
  labs(x = NULL,
       y = NULL,
       title = "The new normal:\nHome deaths increased by a third during the pandemic...",
       subtitle = "...and remain increased in 2023!"
       ) +
  NULL
)

fig_annual_home_deaths %>%
save_output_file(filename = file.path(dir_event, paste0(c("fig_annual_home_deaths", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 10, dpi = 300, units = "cm")
```
